I Compensate, therefore I Am \\ (accurate for DNN verification)

### 2024-01-27 Paper/Paper_noProof.tex
added new files

\newcommand{\toolname}{\ensuremath{\mathsf{Comp}}}

@@

	\section{Value Abstraction for DNN verification}

	In this section, we describe different value (over-)abstractions on $\vz$ that are used by efficient algorithms to certify robustness around an input $\vx$. Over-abstractions of values include all values for $\vz$ in the neighbourhood of $\vx$, and thus a certificate for safety in the over-abstraction is a proof of safety for the original input $\vx$.

	\subsection{The Box and DeepPoly Abstractions}

\begin{figure}[t!]
	\centering
	\begin{tikzpicture}

		\node[circle, draw= black, thick, minimum width = 20,
		minimum height = 20] (input1) {$n_1$};

		\node[circle, draw= purple, thick, minimum width = 20,
		minimum height = 20] (input2) at ($(input1) + (0,-1.5)$) {$n_2$};

		% Hidden layers

		\node (hidden10) at ($(input1) + (2.5,0.6)$) {$0$};

		\node (hidden20) at ($(input1) + (2.5,-1.5-0.6)$) {$0$};

		\node (hidden50) at ($(input1) + (7.5,0.6)$) {$0$};

		\node (hidden60) at ($(input1) + (7.5,-1.5-0.6)$) {$1.5$};

		\node[circle, draw= black, thick, minimum width = 20,
		minimum height = 20] (hidden1) at ($(input1) + (2.5,0)$) {$n_3$};
		\node[circle, draw= black, thick] (hidden2) at ($(input1) + (2.5,-1.5)$) {$n_4$};

		\node[circle, draw= black, thick, minimum width = 20,
		minimum height = 20] (hidden3) at ($(input1) + (5,0)$){$\hat{n}_3$};
		\node[circle, draw= black, thick] (hidden4) at ($(input1) + (5,-1.5)$) {$\hat{n}_4$};

		\node[circle, draw= blue, thick, minimum width = 20,
		minimum height = 20] (hidden5) at ($(input1) + (7.5,0)$){$n_5$};
		\node[circle, draw= black, thick] (hidden6) at ($(input1) + (7.5,-1.5)$) {$n_6$};

		% Output layer
		\node[circle, draw= black, thick, minimum width = 20,
		minimum height = 20] (output1) at ($(input1) + (10,0)$){$\hat{n}_5$};

		\node[circle, draw= black, thick, minimum width = 20,
		minimum height = 20] (output2) at ($(input1) + (10,-1.5)$){$\hat{n}_{6}$};

		% Connections

		\draw[->,thick] ($(input1) + (-1.5,0)$) -- (input1) node[midway, above] {$[-1,1]$};

		\draw[->,thick] ($(input1) + (-1.5,-1.5)$) -- (input2) node[midway, above] {$[-1,1]$};

		\draw[->,thick] (input1) -- (hidden1) node[near start, above] {$1$};
		\draw[->,thick] (input1) -- (hidden2)node[near start, above] {$1$};

		\draw[->,color=green, thick] (input2) -- (hidden1) node[near start, below] {$1$};
		\draw[->,color=red, thick] (input2) -- (hidden2)node[near start, below] {$-1$};

		\draw[->,color=green, thick] (hidden1) -- (hidden3) node[midway, above] {$\max(n_1,0)$};
		\draw[->,color=red, thick] (hidden2) -- (hidden4) node[midway, above] {$\max(n_2,0)$};

		\draw[->,color=green, thick] (hidden3) -- (hidden5) node[near start, above] {$1$};			
		\draw[->,thick] (hidden3) -- (hidden6) node[near start, above] {$0$};

		\draw[->,color=red, thick] (hidden4) -- (hidden5)node[near start, below] {$2$};
		\draw[->,thick] (hidden4) -- (hidden6)node[near start, below] {$1$};

		\draw[->,thick] (hidden5) -- (output1) node[midway, above] {$\max(n_5,0)$};
		\draw[->,thick] (hidden6) -- (output2) node[midway, above] {$\max(n_6,0)$};

	\end{tikzpicture}
	\caption{A DNN example from \cite{kpoly}: every neuron is separated into 2 nodes, $n$ pre- and $\hat{n}$ post-ReLU activation function. The pair $(n_2 n_3 n_5,n_2 n_4 n_5)$ in green and red is compensating (weights of paths are $1,-2$).}
	\label{fig1}
\end{figure}

The idea to perform value abstraction is to compute upper and lower bounds on the values of (some) neurons in the DNN when the inputs are in a given neighbourhood, in order to conclude on the robustness without needing to compute the values exactly for all inputs (in the neighbourhood).

First, remark that weighted sums is a linear function, so it is easy to represent it explicitly. The only problematic part to represent accurately is the ReLU function. While it is fairly simple piecewise linear function with 2 modes ($x<0$ and $x \geq 0$), it is not linear. Considering its iterated impact along the different layers of a ReLU DNN is much more complex than it seems. First, let us remark that it is easy to represent exactly $\ReLU(x)$ in case $x$ is {\em stable}, that is, either it is known to be entirely positive, order
it is known to be entirely negative, as in both case there is only one known linear mode involved. Hence, the sole problem remains how to handle {\em unstable} neurons.

Consider the simpler abstraction, called the {\em Box abstraction} \cite{deeppoly}: inductively compute the bound on each neuron of the next layer independently, by considering the weighted sum over the bounds of the previous layer, then clip the lower bound at 
$\max(0,$ lower bound$)$ to represent the ReLU function, and so on.
For all $i$, let $x_i=\val_{\vx}(n_i)$, with $\vx=(x_1,x_2)$.
For instance, on the DNN example of Fig \ref{fig1}, $x_1,x_2 \in [-1,1]$, then $x_3,x_4 \in [-2,2]$, using ReLU $\hat{x}_3,\hat{x}_4 \in [0,2]$, and $x_5 \in [0,6]$ while $x_6 \in [0,2]$. 
While the bounds for $n_1, \ldots, n_4$ are exact (for all $\alpha$ 
in the range, there is an input $\vy$ such that $\val_{\vy}(n_i)=\alpha$), 
it is no more the case from the next layer (starting in $n_5,n_6$) because of potential 
dependencies between earlier neurons: $n_3$ reaches value $2$ only when $x_1=x_2=2$,
in which case $n_4$ reaches value $Val_{(2,2)}(n_4)=0$. Hence, it is impossible that $x_5=\Val_{\vx}(n_5)$ reaches $6$, as it would require both $n_3$ and $n_4$ to have value $2$.

A still extremely fast algorithm that corrects some of this issue is {\em DeepPoly} \cite{deeppoly}, also rediscovered as the {\em CROWN} algorithm \cite{crown}. Instead of 
hard bounds, for each neuron $n$ of layer $k$, two affine functions (with input at the previous layer $k-1$) are kept as lower bound and upper bound of the value of the node.
Denote $f_i \leq x_i \leq g_i$ with e.g.  
$f_{3}(x_3)=f_4(x_4)=0$ and 
$g_3(x_3) = \frac{x_3+2}{2}$,
$g_4(x_4) = \frac{x_4+2}{2}$. 
With that, we get 
$x_5 \leq g_3(x_3) + 2 g_4(x_4) = \frac{x_3 + 2x_4 + 6}{2} = \frac{3x_1 - x_2 + 6}{2}\leq 5$.

For $[\alpha,\beta]$ bounds on $x_i$, there is one optimal linear function for the upper bound $g_i(x_i)= \beta \frac{x_i-\alpha}{\beta-\alpha}$ for ReLU nodes.
There are two choices for the lower bound: $f^1_i(x_i) = 0$ or $f^2_i(x_i)=x_i$.
In the DeepPoly algorithm, the choice is made by considering $\alpha < 0 < \beta$ (for unstable neurons). If $|\alpha|\geq |\beta|$ then $f_i=f^1_i$ is used, while if $|\beta|>|\alpha|$ then $f_i=f^2_i$ is used. Denote {\em $\overline{\mbox{DeepPoly}}$} the DeepPoly algorithm such that $f^1_i$ is always used, and $f^2_i$ is never used. 
Unlike DeepPoly, {\em $\overline{\mbox{DeepPoly}}$} subsumes the {\em Box abstraction}. Indeed, taking bounds $[-0.2,5]$ on $x_i$, it gives $\ReLU(x_i) \in [0,5]$, 
while the normal DeepPoly algorithm would only allow to conclude that $\ReLU(x_i) \in [-0.2,5]$.

\iffalse
\subsection{PRIMA and $\beta$-CROWN}
\fi

\subsection{MILP and LP encodings for DNNs}

At the other end of the spectrum, we find the Mixed Integer Linear Programming (MILP) value abstraction, which is a priori a complete method (albeit much slower than DeepPoly). 
Consider an unstable neuron $n \in[\alpha,\beta]$. The value $x$ of $\ReLU(n)$ can be encoded exactly in an MILP formula with one integer (actually even binary) variable $a$ valued in ${0,1}$, using 4 constraints \cite{MILP}:
\vspace{-0.1cm}
\begin{align*}
	\hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \beta \cdot a \quad \wedge \quad \hat{x} \leq x-\alpha \cdot (1-a)
\end{align*}

From \cite{MILP}, for all $x \in [\alpha,\beta] \setminus 0$, there is a unique solution $(a,\hat{x})$ satisfying the constraints, and we have $\hat{x}=\ReLU(x)$ (and $a$ is 0 if $x < 0$ and 1 if $x>0$, and can be either for $x=0$). Such an encoding can be used for every (unstable) ReLU node, and optimizing its value allows to certify a given input. 
For Networks with hundreds of nodes or more, the associated MILP formula will have have hundreds of integer variables, and it will usually not be solvable efficiently.

MILP instances can be linearly relaxed into LP over-abstraction, where variables which were integers in $\{0,1\}$ (binary) are relaxed to be real in interval $[0,1]$, using otherwise the same encoding. Solving LP instances being Polynomial time, optimizing it is much more efficient, but at the price of a lack of accuracy (e.g. with not as tight bounds). This is called the {\em LP abstraction}.

As far as we know, no formal result states the relationship between the LP approximation and DeepPoly.
We prove now that the LP abstraction is strictly more accurate than the DeepPoly abstraction, as it actually corresponds to considering in DeepPoly both linear functions lower bounding the ReLU function, $ReLU(x) \geq 0$ and $ReLU(x) \geq x$, while DeepPoly considers exactly one.

\begin{proposition}
	\label{LP}
	Given $x \in [\alpha,\beta]$ with $\alpha < 0 < \beta$, the following two systems of constraints 
	1) for LP and 2) for DeepPoly with both lower bounds are equivalent:
	\vspace{-0.3cm}
	\begin{align*}
		& 1) \quad \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \beta \cdot a \quad \wedge \quad \hat{x} \leq x-\alpha \cdot (1-a), \, a \in [0,1] \\
		\text{and} \quad  & 2)  \quad \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0 \quad \wedge \quad \hat{x} \leq \beta \frac{x-\alpha}{\beta-\alpha}
	\end{align*} 
\end{proposition}

\begin{proof}
	Obviously, the 2 lower bound constraints are the same in 1) and 2).

	It remains to prove that the 2 upper bound constraints of 1) LP corresponds to the 2) DeepPoly linear upper bound. For that, we study the upper bound function in the linear variable $a \in  [0,1]$, with $x \in [\alpha,\beta]$ fixed. We have $\hat{x}$ in 1) is upper bounded by $max_{a \in [0,1]} (min(\beta \cdot a, x - \alpha (1-a)))$, and this bound can be reached. 
	The function $min(\beta \cdot a, x - \alpha (1-a))$ reaches its maximum for 
	$\beta \cdot a = x - \alpha (1-a)$, that is 
	$(\beta - \alpha) a = x - \alpha$, i.e. 
	$ a = \frac{x - \alpha}{\beta-\alpha}$.
	It thus gives an upper bound at 
	$\beta \cdot a = \beta \frac{x - \alpha}{\beta-\alpha}$, which is exactly the upper-bound function used in DeepPoly.
	\hfill $\square$
\end{proof}

>>>>

\newcommand{\toolname}{\CMP}

@@

\input{valabstraction}
